---
title: "1. Detecting bounded transcript-usage episodes"
author: "Robin Gounder"
date: "2026-08-23"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{1. Detecting bounded transcript-usage episodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Differential transcript-usage (DTU) tools estimate evidence for individual
contrasts. In an ordered multi-group experiment, a researcher may instead want
the narrower pattern in which one group separates coherently from at least two
alternatives and subsequently reconverges. Manually collecting significant
pairwise rows does not define the temporal unit, flanks, missing-data behavior,
replicate criterion, or multiplicity family.

`transientDTU` provides that post-inference layer. It does not replace satuRn,
DEXSeq, DRIMSeq, limma, or other DTU engines. It standardizes their signed
effects and adjusted p-values, detects bounded episodes, optionally verifies
replicate separation, annotates reciprocal exchanges, and creates a
deterministic candidate ranking. Its Bioconductor role is interoperability with
`SummarizedExperiment` and existing transcriptomic inference workflows.

### Relationship to existing Bioconductor methods

The packages address complementary layers of an analysis. satuRn, DEXSeq,
DRIMSeq, and limma estimate feature- or transcript-level differential-usage
evidence. Their effects and adjusted p-values can supply the generic pairwise
contract after reshaping. stageR provides stage-wise screening and confirmation
with gene-level false-discovery-rate control for predefined hypotheses; it does
not define bounded diverge-reconverge intervals.

`transientDTU` instead converts ordered, multi-group evidence into explicit
bounded episodes and deterministic candidate rankings. It operates after
upstream inference and does not claim episode-level false-discovery-rate
control.

This separation is deliberate: researchers retain the upstream model suited to
their counts, replication, and experimental design while applying one auditable
temporal decision rule to its results.

## Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("transientDTU")
```

```{r load}
library(transientDTU)
data("transientExampleSE")
data("transientExamplePairwise")
example_metadata <- S4Vectors::metadata(transientExampleSE)
```

## Generic pairwise input

The atomic input row is one feature, gene, focal group, comparator group, and
stage. `effect` must be the signed quantity

\[
\mathrm{usage}_{focal} - \mathrm{usage}_{comparator}.
\]

An adjusted p-value belongs to the complete upstream testing family. Optional
gene-level adjusted p-values can impose a second screen.

```{r input}
head(transientExamplePairwise)
```

Column names are configurable, so the table can originate from any upstream
method. Duplicate atomic contrasts are rejected rather than silently reduced.

Run a preflight audit before discovery. It distinguishes blocking structural
errors from contrast-coverage or replicate warnings, while explicitly reminding
the user which inferential properties cannot be checked from a result table.

```{r audit}
input_audit <- auditTransientInput(
    transientExamplePairwise,
    example_metadata$stage_order,
    usage = transientExampleSE
)
input_audit
head(input_audit$coverage)
```

## Standardize stage evidence

```{r stage}
stage <- makeStageDTU(
    transientExamplePairwise,
    stage_order = example_metadata$stage_order,
    gene_name_col = "gene_name",
    gene_q_col = "gene_q"
)
stage
```

For each feature, focal group, and stage, this table records the weakest
component q-value, minimum and maximum absolute effect, effect direction,
comparator spread, and whether every expected comparator was observed.

## Detect and annotate episodes

```{r detect}
episodes <- detectEpisodes(
    stage,
    q_threshold = 0.05,
    effect_threshold = 0.10,
    comparator_tolerance = 0.10,
    flank_tolerance = 0.10,
    gene_q_threshold = 0.05
)
episodes
```

Consecutive equal non-zero states collapse to one interval. The immediate stage
before and after the interval must be observed and must meet the flank
tolerance; boundary runs are not called.

Duration and flank depth are explicit rather than inferred from labels. Numeric
coordinates annotate irregular spacing, while episode adjacency remains the
declared stage order.

```{r episode-controls}
stage_coordinates <- c(0, 1, 2, 4, 8, 16, 32, 64)
controlled <- detectEpisodes(
    stage,
    min_episode_stages = 1,
    max_episode_stages = 2,
    flank_width = 1,
    stage_coordinates = stage_coordinates
)
controlled[, c(
    "start_stage", "end_stage", "n_stages", "coordinate_span"
)]
```

The legacy-compatible default uses all available finite flank summaries and
records `n_finite_flanks` and `flanking_complete`. A stricter sensitivity is:

```{r complete-flanks}
complete_flanks <- detectEpisodes(
    stage,
    gene_q_threshold = 0.05,
    flank_missing = "complete"
)
c(available = nrow(episodes), complete = nrow(complete_flanks))
```

Replicate separation verifies that every focal replicate is beyond every
non-focal replicate in the required direction at every episode stage.

```{r replicate}
episodes <- checkReplicateSeparation(
    episodes,
    transientExampleSE,
    keep = "consistent"
)
episodes <- annotateReciprocal(episodes)
rankCandidates(
    episodes,
    n = 6,
    reciprocal_only = TRUE,
    replicate_consistent_only = TRUE
)
```

Complete separation is intentionally stringent. For larger cohorts where one
outlier should not define the descriptive check, request a quantile or median
summary and state that choice in the report. Missing values may be omitted only
with an explicit minimum finite replicate count.

```{r robust-replicates}
quantile_checked <- checkReplicateSeparation(
    detectEpisodes(stage),
    transientExampleSE,
    replicate_method = "quantile",
    quantile_probability = 0.10,
    missing_values = "omit",
    min_replicates = 2
)
table(quantile_checked$replicate_consistent, useNA = "ifany")
```

## Two-group and targeted-focal designs

The motivating study required separation from at least two alternatives. A
two-group study is also valid when the scientific question only requires one
comparator; set `min_comparators = 1` in the audit and workflow. A table may
contain only the focal groups intended for scanning; the audit infers those
from the focal column unless `expected_focal_groups` is supplied.

```{r two-groups}
two_group <- simulateTransientDTU(
    n_genes = 8, groups = c("treated", "control"), seed = 902
)
two_group_audit <- auditTransientInput(
    two_group$pairwise,
    two_group$stage_order,
    min_comparators = 1
)
two_group_fit <- runTransientDTU(
    two_group$pairwise,
    two_group$stage_order,
    min_comparators = 1,
    reciprocal_only = FALSE
)
c(status = two_group_audit$status, episodes = nrow(episodeTable(two_group_fit)))
```

## Complete workflow

`runTransientDTU()` performs the same sequence and returns a validated result
object.

```{r workflow}
result <- runTransientDTU(
    transientExamplePairwise,
    stage_order = example_metadata$stage_order,
    usage = transientExampleSE,
    gene_name_col = "gene_name",
    gene_q_col = "gene_q",
    gene_q_threshold = 0.05,
    panel_size = 6
)
result
candidateTable(result)
diagnosticTable(result)
```

```{r trajectory, fig.width=7, fig.height=4}
example_feature <- example_metadata$truth$feature_id[[1]]
plotTransientTrajectory(
    result,
    feature_id = example_feature,
    focal_group = example_metadata$groups[[1]]
)
```

## SummarizedExperiment interoperability

```{r se}
transientExampleSE
SummarizedExperiment::assayNames(transientExampleSE)
head(SummarizedExperiment::rowData(transientExampleSE))
head(SummarizedExperiment::colData(transientExampleSE))
checked_from_se <- checkReplicateSeparation(
    detectEpisodes(stage), transientExampleSE, keep = "consistent"
)
nrow(checked_from_se)
```

Users starting from proportions can optionally obtain generic pairwise evidence
with `fitPairwiseUsage()`. This convenience adapter uses a saturated limma
cell-means model. A purpose-built DTU model remains preferable when its
assumptions match the data.

## Joint threshold stability

```{r stability}
stability <- thresholdStability(
    stage,
    q_threshold = c(0.01, 0.05),
    effect_threshold = c(0.10, 0.15),
    comparator_tolerance = c(0.10, 0.15),
    flank_tolerance = c(0.10, 0.15),
    gene_q_threshold = list(NULL, 0.05),
    usage = transientExampleSE,
    replicate_action = "filter"
)
head(stability$summary)
stability$selection_frequency
```

The Cartesian grid is deliberate. Varying one threshold at a time can conceal
interactions between evidence, effect, comparator, and flank requirements.

## Interpretation boundary

An episode is a candidate satisfying a decision rule, not an independently
validated developmental switch. Component-test adjustment does not imply
episode-level FDR control. Bulk composition, transcript identifiability,
post-selection inference, technical batches, and external validation must be
addressed in the scientific application.

## References

- Anders S, Reyes A, Huber W (2012). Detecting differential usage of exons
    from RNA-seq data. *Genome Research*, 22, 2008--2017.
    doi:10.1101/gr.133744.111.
- Gilis J, Vitting-Seerup K, Van den Berge K, Clement L (2021). satuRn:
    scalable analysis of differential transcript usage for bulk and single-cell
    RNA-sequencing applications. *F1000Research*, 10, 374.
    doi:10.12688/f1000research.51749.2.
- Nowicka M, Robinson MD (2016). DRIMSeq: a Dirichlet-multinomial framework for
    multivariate count outcomes in genomics. *F1000Research*, 5, 1356.
    doi:10.12688/f1000research.8900.2.
- Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W, Smyth GK (2015). limma
    powers differential expression analyses for RNA-sequencing and microarray
    studies. *Nucleic Acids Research*, 43, e47. doi:10.1093/nar/gkv007.
- Van den Berge K, Soneson C, Robinson MD, Clement L (2017). stageR: a general
    stage-wise method for controlling the gene-level false discovery rate in
    differential expression and differential transcript usage.
    *Genome Biology*, 18, 151. doi:10.1186/s13059-017-1277-0.

## Session information

```{r session-info}
sessionInfo()
```
