---
title: "transientDTU: detecting bounded transcript-usage episodes"
author: "Robin Gounder"
date: "2026-08-23"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{1. Detecting bounded transcript-usage episodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Differential transcript-usage (DTU) tools estimate evidence for individual
contrasts. In an ordered multi-group experiment, a researcher may instead want
the narrower pattern in which one group separates coherently from at least two
alternatives and subsequently reconverges. Manually collecting significant
pairwise rows does not define the temporal unit, flanks, missing-data behavior,
replicate criterion, or multiplicity family.

`transientDTU` provides that post-inference layer. It does not replace satuRn,
DEXSeq, DRIMSeq, limma, or other DTU engines. It standardizes their signed
effects and adjusted p-values, detects bounded episodes, optionally verifies
replicate separation, annotates reciprocal exchanges, and creates a
deterministic candidate ranking. Its Bioconductor role is interoperability with
`SummarizedExperiment` and existing transcriptomic inference workflows.

## Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("transientDTU")
```

```{r load}
library(transientDTU)
data("transientExample")
```

## Generic pairwise input

The atomic input row is one feature, gene, focal group, comparator group, and
stage. `effect` must be the signed quantity

\[
\mathrm{usage}_{focal} - \mathrm{usage}_{comparator}.
\]

An adjusted p-value belongs to the complete upstream testing family. Optional
gene-level adjusted p-values can impose a second screen.

```{r input}
head(transientExample$pairwise)
```

Column names are configurable, so the table can originate from any upstream
method. Duplicate atomic contrasts are rejected rather than silently reduced.

## Standardize stage evidence

```{r stage}
stage <- makeStageDTU(
    transientExample$pairwise,
    stage_order = transientExample$stage_order,
    gene_name_col = "gene_name",
    gene_q_col = "gene_q"
)
stage
```

For each feature, focal group, and stage, this table records the weakest
component q-value, minimum and maximum absolute effect, effect direction,
comparator spread, and whether every expected comparator was observed.

## Detect and annotate episodes

```{r detect}
episodes <- detectEpisodes(
    stage,
    q_threshold = 0.05,
    effect_threshold = 0.10,
    comparator_tolerance = 0.10,
    flank_tolerance = 0.10,
    gene_q_threshold = 0.05
)
episodes
```

Consecutive equal non-zero states collapse to one interval. The immediate stage
before and after the interval must be observed and must meet the flank
tolerance; boundary runs are not called.

The legacy-compatible default uses all available finite flank summaries and
records `n_finite_flanks` and `flanking_complete`. A stricter sensitivity is:

```{r complete-flanks}
complete_flanks <- detectEpisodes(
    stage,
    gene_q_threshold = 0.05,
    flank_missing = "complete"
)
c(available = nrow(episodes), complete = nrow(complete_flanks))
```

Replicate separation verifies that every focal replicate is beyond every
non-focal replicate in the required direction at every episode stage.

```{r replicate}
episodes <- checkReplicateSeparation(
    episodes,
    transientExample$usage,
    transientExample$col_data,
    keep = "consistent"
)
episodes <- annotateReciprocal(episodes)
rankCandidates(
    episodes,
    n = 6,
    reciprocal_only = TRUE,
    replicate_consistent_only = TRUE
)
```

## Complete workflow

`runTransientDTU()` performs the same sequence and returns a validated result
object.

```{r workflow}
result <- runTransientDTU(
    transientExample$pairwise,
    stage_order = transientExample$stage_order,
    usage = transientExample$usage,
    col_data = transientExample$col_data,
    gene_name_col = "gene_name",
    gene_q_col = "gene_q",
    gene_q_threshold = 0.05,
    panel_size = 6
)
result
candidateTable(result)
diagnosticTable(result)
```

```{r trajectory, fig.width=7, fig.height=4}
example_feature <- transientExample$truth$feature_id[[1]]
plotTransientTrajectory(
    result,
    feature_id = example_feature,
    focal_group = transientExample$groups[[1]]
)
```

## SummarizedExperiment interoperability

```{r se}
se <- SummarizedExperiment::SummarizedExperiment(
    assays = list(usage = transientExample$usage),
    colData = S4Vectors::DataFrame(transientExample$col_data),
    rowData = S4Vectors::DataFrame(
        gene_id = sub("_tx[12]$", "", rownames(transientExample$usage))
    )
)
checked_from_se <- checkReplicateSeparation(
    detectEpisodes(stage), se, keep = "consistent"
)
nrow(checked_from_se)
```

Users starting from proportions can optionally obtain generic pairwise evidence
with `fitPairwiseUsage()`. This convenience adapter uses a saturated limma
cell-means model. A purpose-built DTU model remains preferable when its
assumptions match the data.

## Joint threshold stability

```{r stability}
stability <- thresholdStability(
    stage,
    q_threshold = c(0.01, 0.05),
    effect_threshold = c(0.10, 0.15),
    comparator_tolerance = c(0.10, 0.15),
    flank_tolerance = c(0.10, 0.15),
    gene_q_threshold = list(NULL, 0.05),
    usage = transientExample$usage,
    col_data = transientExample$col_data,
    replicate_action = "filter"
)
head(stability$summary)
stability$selection_frequency
```

The Cartesian grid is deliberate. Varying one threshold at a time can conceal
interactions between evidence, effect, comparator, and flank requirements.

## Interpretation boundary

An episode is a candidate satisfying a decision rule, not an independently
validated developmental switch. Component-test adjustment does not imply
episode-level FDR control. Bulk composition, transcript identifiability,
post-selection inference, technical batches, and external validation must be
addressed in the scientific application.

## Session information

```{r session-info}
sessionInfo()
```
