---
title: "Validating the transientDTU decision layer"
author: "Robin Gounder"
date: "2026-08-23"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2. Validating the transientDTU decision layer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

A reusable post-inference framework needs operating-characteristic evidence,
not only a successful application. This vignette demonstrates known-truth
simulation and joint sensitivity analysis for `transientDTU`. The simulator is
intentionally scoped to the decision layer: it does not claim to reproduce
read assignment, count dispersion, or every upstream DTU model.

```{r load}
library(transientDTU)
```

## Known trajectory classes

```{r simulate}
simulated <- simulateTransientDTU(
    n_genes = 24,
    n_stages = 8,
    n_replicates = 2,
    signal = 0.30,
    noise_sd = 0.015,
    seed = 901
)
table(simulated$gene_truth$trajectory)
```

Null genes have no systematic group effect. Monotonic signals do not
reconverge. Persistent signals reach the final boundary. Transient signals
separate at one interior stage and have observed baseline flanks. Every gene
has two complementary features, making true transient genes reciprocal.

```{r one-fit}
fit <- runTransientDTU(
    simulated$pairwise,
    simulated$stage_order,
    usage = simulated$usage,
    col_data = simulated$col_data,
    gene_name_col = "gene_name",
    gene_q_col = "gene_q",
    panel_size = Inf
)
diagnosticTable(fit)
```

## Repeated known-truth benchmark

```{r benchmark}
benchmark <- benchmarkTransientDTU(
    n_simulations = 3,
    simulation_args = list(
        n_genes = 20,
        signal = 0.30,
        noise_sd = 0.015
    ),
    detector_args = list(
        gene_name_col = "gene_name",
        gene_q_col = "gene_q"
    ),
    seed = 1000
)
benchmark$results
benchmark$summary
```

Precision is undefined in a run with no detected episodes and is therefore
summarized with missing values removed. `any_false_positive` estimates the
simulation-level probability of at least one false episode; `null_gene_fpr`
reports the fraction of null genes with any detected episode.

## Stress dimensions

Researchers should vary at least:

- signal magnitude and replicate noise;
- number of replicates;
- one-stage versus multi-stage true episodes;
- missing atomic contrasts;
- adjustment method and testing-family definition; and
- all detector thresholds jointly.

```{r missingness}
missing <- benchmarkTransientDTU(
    n_simulations = 2,
    simulation_args = list(
        n_genes = 16,
        missing_fraction = 0.02
    ),
    detector_args = list(
        gene_name_col = "gene_name",
        gene_q_col = "gene_q"
    ),
    seed = 1200
)
missing$summary
```

## Scope of the benchmark

The simulator's p-values use the known generating noise and global adjustment.
They isolate the detector's behavior after inference. A publication claiming
performance for a particular quantifier and DTU engine should additionally
simulate or resample at the count/read-assignment level and apply that complete
pipeline before `transientDTU`.

## Session information

```{r session-info}
sessionInfo()
```
